Phylogenetics and the human microbiome.

### 2014-06-03 reviews/response.txt

> 1) Pg. 3: “It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization” – This suggests that molecular sequences have traditionally been characterized via morphological or phenotypic means. I suppose that’s indirectly true, but I don’t think it is the best way to describe it.

> 2) Pg. 10: “Inference of this history is complicated by the fact that any tree-like signal in the overall evolution of bacteria is restricted to a small set of so-called “core” genes” – Would it be more accurate to write that tree-like signal regarding the deep relationships of bacteria is restricted to a small set of so-called “core” genes?  I suspect all genes carry some tree-like signal, even in microbes, though some have been subject to much more lateral gene transfer than others.

> 3) Pg. 14: “first with a Dirichlet multinomial mixtures on 3-mer profiles” – Should this just be “mixture”?

> 4) Pg. 15: “However, whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate” – This requires some modification.  Maybe add “one” after “every”.

> 5) Pg. 16: “PD under random sub-sampling, which is often applied to enable comparison be- tween samples of different sequencing depths and called “rarefaction curves.”” – Would it be preferable to write that the random subsamples are described (or depicted) with rarefaction curves?

> 6) Pg. 23: “The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history…” – “Begging the question” means something else in philosophical circles. You should probably just use “raises the question” here.

Your revision must comply with our instructions to authors (in this letter, with additional instructions at http://www.oxfordjournals.org/our_journals/sysbio/for_authors). This applies even if your original submission deviated from Systematic Biology style and corrections were not included in the notes from the editors or reviewers. Therefore, READ AND APPLY OUR INSTRUCTIONS BEFORE SUBMITTING A REVISION. Failure to do so will result in significantly delayed processing of your paper. If anything remains unclear after reading the instructions, use recent issues of the journal to find examples. If necessary, feel free to contact the Managing Editor at sysbio.editorialoffice@oup.com for help.

After you've uploaded your revision, carefully view the Manuscript Central version to verify that all figures and other files display correctly, and that you’ve followed all of our author instructions.

The cost of printed color figures is $600 each. Authors are normally expected to cover this cost, but we do have limited funds available for authors who are SSB members, and cannot pay the full amount. If you feel your circumstances create an unusual financial need, please explain in your cover letter. The maximum allowance is one color figure per paper.
All color costs can be avoided if you decide to have the figures printed in black and white, but be shown in color in the online version of the paper. If you do this, the captions must be worded such that they are appropriate for both situations (e.g., descriptions should not name colors), because the captions in print and online will be identical even if color is used online only. Do not submit two versions of any figure. Instead, make sure the color figure is also easily legible in grayscale, then submit only the color version. In your cover letter, please make your intentions regarding the use of color clear.

Journal Covers: Please consider submitting a suggested cover image. They can be illustrations of theory, photos of organisms, a combination of the two, or alternatives. These can be uploaded with a revision of the manuscript or may be sent later. Only images for which you have copyright permission or that are not under copyright may be submitted. The color figure fee does not apply to images chosen by the Editor to be on the cover of the issue. If you have questions about possible cover designs, please email sysbio.editorialoffice@oup.com. We encourage and greatly appreciate cover image suggestions.

Make sure all section headings conform to Systematic Biology style for first, second and third levels. Use of incorrect styles is potentially confusing and, in any case, is likely to delay processing of the manuscript.

Tables should have a single-sentence informative title above the table, with any other descriptive information located below the table in the form of notes and/or specific footnotes.

When references are grouped together in parentheses in the text, they should be listed in ascending chronological order. Multiple references in a single year should be alphabetized.

All funding used for this work should be listed in a "Funding" section preceding the Acknowledgements. Please give the full official name of each funding body.

AE
--

> Thanks for submitting the revised version of your review paper
> “Phylogenetics and the human microbiome” to Systematic Biology
> (special online issue, after MCEB2013). It has been read by the same
> three referees as the original submission: Alexis Stamatakis, an
> anonymous referee (2), and Frédéric Delsuc. All three recommend
> acceptance with minor or no change. Actually, both of them (Alexis and
> Fred) recommend to publish the paper as is, and referee 2 has only
> minor comments on figure 2. I read the paper and reach the same
> conclusion. My (minor) comments are provided below. I believe that you
> have done a great job in putting all of these works and references
> together, in a very comprehensive way, with highly motivating
> perspectives for future research in phylogenetics. I’m thus happy to
> recommend acceptance with quite minor changes (mostly in figures 2 and
> 3).

> 9 (Kuczynski et al. 2010). However

> P12 “Several groups are continually revising taxonomy...” and P13 “An
> obvious application of phylogenetics is to perform taxonomic
> classification...” We have here two long and technical paragraphs
> (e.g. F-measure, precision and recall, but what is classified and
> how?). You should check this again for clarity and to ensure
> self-contentedness.

> P16 connecting taxa in the set (Faith 1992).

> Fig 1 The "graphical formulae" [l(segments) A(segments)] at the bottom
> of the figure do not help. This can be better described by
> checking-improving the figure legend, notably "the abundance of the
> taxa on either side of the branch" (how is it computed? give an
> example if there are several options)

> Fig 2 Your two cases are extreme. Having one (or a few) shared (brown)
> edges in the right tree, and one unshared (blue or green branch) edge
> in the left tree would help. See also comments of referee 2.

> Fig 3 and its legend, as well as the corresponding text P17-18 are
> quite unclear to me.... It would be interesting to have more
> explanations, both on the math side and on the application to sequence
> reads.

Reviewer: 1
-----------

> The author has addressed all of my minor suggestions for revision and
> has also fully addressed the suggestions made by the other reviewers.

Reviewer: 2
-----------

> As noted in my first review, I like the idea of this manuscript,and
> much of the content.  The revised version is very close to the first
> version, so I assume the author is happy with it.  I certainly
> appreciated the diff.pdf.

> I am still not clear how the author chooses what needs technical
> discussion and what does not, and more technical detail has been added
> to this version.  However other, perhaps more technically-inclined
> reviewers, did not really pick this up, so, in the interests of
> efficiency, I won't belabour the point. The author must have a clear
> view who the intended audience is.

> I will say that, while figure 2 does a fairly good job of illustrating
> the critical UniFrac method (though the common brown internal edges
> could be named explicitly in the caption), the definition in the text
> does not: "The name UniFrac is a contraction of “Unique Fraction,”
> which refers to the fact that the original UniFrac definition compares
> the fraction of edges that connect only tips from one sample via the
> shortest path rather than edges that connect between samples (Fig. 2;
> Lozupone and Knight 2005)" makes very little sense to me, particularly
> "edges between samples". Could this be reworded in a more intuitive
> way that refers explicitly to the tree(s) being inferred and the tips
> used to infer the trees?  So, for instance, does one infer one tree
> per sample and then one tree of all samples?

Reviewer: 3
-----------

> I went through both the revised manuscript and the author’s response
> letter. I think that the author has made significant efforts to
> address all editorial and referee comments, including my own concerns.
> I would therefore recommend the acceptation of the revised manuscript
> basically as is.

### 2014-06-03 sbreview.tex
Doing the easy changes.

Where Woese and colleagues labored over digestion and gel electrophoresis to infer sequences, modern researchers have the luxury of high throughput sequencing.
This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
This has led to extensive parallelization, most recently by sequencing dozens of samples at a time on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
This brings up the question of how many sequences are needed to characterize the microbial diversity of a given environment.
To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however to compare more similar samples deeper sequencing is required.
In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep[e.g.][]{caporaso2011moving}.

@@

There are many OTU inference methods with various speeds and strategies.
Some methods proceed through a list of sequences and progressively add each either to an existing cluster or start a new cluster with that sequence, such as CD-HIT \citep{li2006cdhit} and USEARCH \citep{edgar2010usearch}, while UPARSE \citep{edgar2013uparse} uses a similar strategy while also attempting to correct for sequencing error.
\citet{white2010alignment} show that different ways of doing this genre of heuristic clustering can result in very different results.
\citet{Cai2011-bi} perform highly efficient clustering using a pseudometric based partition tree, which can be thought of as a hierarchical clustering tree with a fixed set of internal node heights.
\citet{navlakha2009finding} take a semi-supervised approach in that input includes sequences along with a subset of sequences that are equipped with taxonomic classifications; the algorithm then groups all sequences (many in novel clusters) into clusters that have similar properties as the example taxonomic clusters.
\citet{Wang2013-ol} optimize a criterion of cluster modularity.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a relatively large number of sequencing reads and so can be used as an OTU clustering method.

@@

Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
For example, the most commonly used index is the Simpson index \citep{simpson1949measurement}, which is simply the sum of the squared frequencies of the OTUs.
The drawback of applying such a diversity metric to a collection of OTUs is that the large-scale structure of the diversity is lost, such that two closely related OTUs contribute as much to the diversity measure as two distant ones.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used.
However, whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD.

@@

The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
This suggests a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of microbes that allows us to best perform proxy whole-genome comparison?
This representation could be simple.
For example, one of the results of \citet{zaneveld2010ribosomal} is that 16S correlates better with gene repertoire in some taxonomic groups than others.
If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more precise comparison?
If we allow an arbitrary ``hidden'' object, what such object would perform best?
\citet{Parks2012-os} have expanded the range of choices by defining community comparison metrics on phylogenetic split systems.
An alternative would be to use collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[e.g.][]{szollHosi2013efficient,szollHosi2013lateral}.

>>>>

Where Woese and colleagues labored over digestion and gel electrophoresis to infer sequences, modern researchers have the luxury of high throughput sequencing.
This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
This has led to extensive parallelization, most recently by sequencing dozens of samples at a time on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
This brings up the question of how many sequences are needed to characterize the microbial diversity of a given environment.
To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial}, however, to compare more similar samples deeper sequencing is required.
In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep[e.g.][]{caporaso2011moving}.

@@

There are many OTU inference methods with various speeds and strategies.
Some methods proceed through a list of sequences and progressively add each either to an existing cluster or start a new cluster with that sequence, such as CD-HIT \citep{li2006cdhit} and USEARCH \citep{edgar2010usearch}, while UPARSE \citep{edgar2013uparse} uses a similar strategy while also attempting to correct for sequencing error.
\citet{white2010alignment} show that different ways of doing this genre of heuristic clustering can result in very different results.
\citet{Cai2011-bi} perform highly efficient clustering using a pseudometric based partition tree, which can be thought of as a hierarchical clustering tree with a fixed set of internal node heights.
\citet{navlakha2009finding} take a semi-supervised approach in that input includes sequences along with a subset of sequences that are equipped with taxonomic classifications; the algorithm then groups all sequences (many in novel clusters) into clusters that have similar properties as the example taxonomic clusters.
\citet{Wang2013-ol} optimize a criterion of cluster modularity.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixture on 4-mer profiles, and then a minimum description length criterion.
\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a relatively large number of sequencing reads and so can be used as an OTU clustering method.

@@

Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
For example, the most commonly used index is the Simpson index \citep{simpson1949measurement}, which is simply the sum of the squared frequencies of the OTUs.
The drawback of applying such a diversity metric to a collection of OTUs is that the large-scale structure of the diversity is lost, such that two closely related OTUs contribute as much to the diversity measure as two distant ones.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used.
However, whereas just about every one of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD.

@@

The apparent success of 16S-tree-based comparisons raises the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
This suggests a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of microbes that allows us to best perform proxy whole-genome comparison?
This representation could be simple.
For example, one of the results of \citet{zaneveld2010ribosomal} is that 16S correlates better with gene repertoire in some taxonomic groups than others.
If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more precise comparison?
If we allow an arbitrary ``hidden'' object, what such object would perform best?
\citet{Parks2012-os} have expanded the range of choices by defining community comparison metrics on phylogenetic split systems.
An alternative would be to use collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[e.g.][]{szollHosi2013efficient,szollHosi2013lateral}.
